---
title: "betteR"
subtitle: "Learnings from an everyday pursuit of wanting to write better R code"
author: "Kaushik Lakshman"
output:
  xaringan::moon_reader:
    css: [default, rladies, rladies-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# About me & today

- Lead the Data Science team @ Betfair 

- Massive fan of R & the community around it

- Always learning & always looking to improve

- Want to share my experiences and learnings in the pursuit of the above motto

- Typically look back at my code from 6 months ago and simultaneously feel both 

  - Disgusted at the old code
  
  - Very happy about how far things have come

---

# Make your R code spark joy!
.center[
<img src = "imgs/marie_kondo_1.png">
]
---

class: inverse, middle, center

# Basics

---


# Readable code

- Comments

  - Comment as much as you can 
  
  - Think about others who may read or run your code 
  
  - If you don't care about others, think about future you, she/he will definitely appreciate it

- Have a consistent stlye guide 

  - Tidyverse style guide
  
  - Google style guide 
  
  - Whatever works for you, but consistent 


---

# Tidyverse style guide

- Names 

  - Meaningful names for variables, functions, files 
  
  - Recommended that object names be nouns and function names be verbs 
  
  - Underscores instead of dots 
  
```{r, eval = FALSE, echo=TRUE}
## Sparks joy 
weather_data 
weather_analysis.R
clean_weather_data()

## Does not spark joy
w_df 
wAnalysis.R
weather.cleaner()
```

---

# Scoping

- Functions can have the same name, but do different things, and come from different packages 

- Which function gets used when? 

- Always helpful to reference the package from which you are using the function 

```{r, eval=FALSE, echo=TRUE}
## Helpful to call function like this
package::function()
  
## Examples
glue::glue_sql()
logger::log_info()
```

- In addition to proper scoping, this method also helps in failing loudly on errors, making debugging easier 
---

# Environment

- Many of my scripts used to start like this 

```{r, eval=FALSE, echo=TRUE}
setwd("C:/some_folder/some_subfolder")
rm(list = ls())
```

- Working directory

  - Specific to your computer

  - What happenes if your teammate wants to run this? What happens if you want to run the same analysis on another machine?
  
  - Solution -> R Studio Projects; Here package
  
- Clean environment 

  - `rm` only removes objects from your environment, not loaded packages
  
  - Solution -> Restart R (disable save environment option) for a clean environment


---

# Projects & here

- Rstudio offers Projects that lets you organise logical bits of code in the same place 

- Key components of projects are 

  - New R session started every time you open a project 
  - Working directory is set by default to project directory 
  - Git integration 
  
- Along with Projects the `here` package is super useful for relative path definitons 

- `here` creates a `.here` file which sits by default in your project directory and you can use the functions from the package to refer to other files in the directory 

```{r eval=FALSE, echo=TRUE}
read_csv(here::here("folder", "subfolder", "file_that_i_want_to_read.csv"))
```

- Advantages of using Projects & here - things are reproducible across environments and across users

---

class: inverse, middle, center

# Don't Repeat Yourself 

---

# DRY Principle

- If you write the same code three times, put it in a function 

- If you write the same function three times put it in a package

- Reduce, Reuse and Refactor & make Captain rPlanet happy

.center[
<img src = "imgs/captain_planet_1.png">
]

---

# Functions 

- Why functions? 
  
  - Modularity 
  - Clean global environments
  - Easier to make changes
  - Easier to test

```{r, eval=FALSE, echo=TRUE}
calculate_something_last_x_days <-
  function(.raw_data,
           .last_x_days) {
    
    ## Filter based on parameter & then calculate
    .raw_data %>%
      filter(some_date_filter >= .last_x_days) %>%
      some_calculation()
    
  }
```

- Style guide 
  
  - Parameters starting with a .
  - No explicit return

---

# Tidy Eval

- Until some time ago functions had a limited scope
  - Could do anything with data inside a data frame in functions
  - Far more complex to manipulate columns inside functions 
  
```{r, eval=FALSE, echo=TRUE}
get_max_by <- function(.data, .variable, .by) {
  .data %>%
    group_by(.by) %>%
    summarise(maximum_value = max(.variable, na.rm = TRUE))
}

## This will error out
starwars %>% get_max_by(mass, .by = gender)
```

- Initially solved by `quo`, `enquo`, and `sym` pattern; works but its complex 

```{r, eval=FALSE, echo=TRUE}
get_max_by <- function(.data, .variable, .by) {
  data %>%
    group_by(!!enquo(.by)) %>%
    summarise(maximum = max(!!enquo(.variable), na.rm = TRUE))
}

## This will work
starwars %>% get_max_by(mass, .by = gender)
```

---

# Tidy Eval

- Since `rlang 0.4.0` there is an alternate, simpler pattern for such code 

```{r eval=FALSE, echo=TRUE}
get_max_by <- function(.data, .variable, .by) {
  data %>%
    group_by({{ .by }}) %>%
    summarise(maximum = max({{ .var }}, na.rm = TRUE))
}

## This will work
starwars %>% get_max_by(height, .by = gender)
```

---

# Packages

- Writing a package is not as daunting as it seems

- Your package can just be a collection of functions you use often 

- Your package can be just for yourself, and not exposed to the world 

- Your package does not have to end up on CRAN 

- Package writing requires no extra skills - If you can write R code, you can write a package 

- Only prerequisite is the ability to install and use the packages `usethis` and `devtools` 

- Some great getting started resouces

  - Old but good intro by Hillary Parker <a href= "https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/">  Writing an R package from scratch </a>
  
  - Melbourne's own Earo Wang did a presentation at R-Ladies Melbourne called <a href = "https://github.com/earowang/rladies-pkg"> Everything you need to create your package </a>

---

# Packages 

- The package making toolkit contains a few packages. Not all are necessarily but below is a list. 

  - `available` to check for package names 
  
  - `usethis` to automating package set up and other related things 
  
  - `devtools` to automate auxillary things about package writing like building documentation etc 
  
  - `roxygen2` to convert your documentation notes into help files
  
  - `testthat` to write unit tests 
  
  - `pkgdown` to build a Rmarkdown based website for your package

---

# Packages 

- Use `usethis::create_package()` or R-Studio's New project -> R Package flow to get started 

- You will get a basic file structure created that has the below components 

  - Folder called `R/` that contains all the functions in your package 
  
  - Folder called `man/` that contains documentation for all the functions in your package
  
  - `DESCRIPTION` for metadata that you can edit with package details and `NAMESPACE` for details about how R deals with other packages
  
- Use `usethis::use_roxygen_md()` to let your package know that you will be using Roxygen to build your documentation 

- You are ready to write functions into your R package! 
---

# Packages

- Use `usethis::use_r()` or Create a new R file in the `R/` directory 

- Insert Roxygen skeleton into the file and edit documentation details 

  - `@param` for parameters
  - `@return` for return type
  - `@export` if you want to make the function available for the package users
  - `@import` to let the function know about dependencies
  - `@examples` to show the user how to use the function

```{r, echo=TRUE, eval=FALSE}
#' This function makes you a coffee based on what you pass as parameters
#'
#' @param .type One of Flat White, Latte or Cappucino 
#' @param .milk data frame of the type of milk you would like 
#' @param .sugar Logical, defaults to FALSE 
#' @param .beans data frame of coffee beans 
#'
#' @return Coffee object
#' @export
#' @import tidycoffee
#'
#' @examples
make_coffee(.type = "Latte", .milk = soy_milk,
            .sugar = FALSE, .beans = guatemala_beans)
```

---

# Packages

- Couple of differences writing functions in packages as opposed to normal 

  - No `setwd()` or any such 
  - No `library()` - use `@import` or `usethis::use_package()` instead 
  - No `options()` or `source()` - you can use other functions in the package
  
- Write your function as normal 

```{r, eval=FALSE, echo=TRUE}
make_coffee <- function(.type, .milk, .sugar = FALSE, .beans) {
  
  feed_into_coffee_machine(
    .beans %>% grind_beans(),
    .milk %>% boil_milk(),
    .sugar %>% add_sugar(),
    .type %>% calculate_proportions(.type)
  )
  
}
```

- Finally run `devtools::document()` which will build documentation and then build package

- That's it! A couple of extra things to keep in mind and you are now equipped to build your own package!
---

## Captain rPlanet is happy!

.center[
<img src = "imgs/captain_planet_2.png">
]

---

class: inverse, middle, center

# Uncategorised stuff

---

# Purrr

- For loops have a bad reputation in R that they are slow. In reality that's not always the case, it is because it is harder for humans to read for loop code 

- Alternative is to use split-apply-combine methods like the functions in the `purrr` package. `purrr` has a lot of amazing functions but the best place to start is the `map_*` family of functions
  
  - `map` takes in a single vector parameter, iterates over it, and applies a function to each element
  
  - `map` returns a list with one output for each iteration 
  
  - `map` has extra functions that help you combine your result 
      
      - `map_chr` if your result is a character vector 
      
      - `map_dbl` if your result is a numeric vector
      
      - `map_dfr` if your result is a data frame that has to be bound by rows 


---

# Purrr 

```{r, echo=TRUE, eval=FALSE} 
map(
  .x = c("Melbourne", "Sydney", "Brisbane", "Adelaide", "Perth"),
  .f = function(sales, x) {
    write_csv(sales %>%
                filter(city == x),
              glue::glue("sales_{x}.csv"))
  }
)
```

- `map2_*` functions are simlar to `map` except they cycle over two vectors `.x` and `.y`

- If you want to iterate over more than 2 vectors use the `pmap_*` function 

- `purrr` also has some new and very cool functions called `safely`, `possibly`, and `quietly` that help you capture side effects and errors in a more flexible manner as you `map` over things. 

```{r, echo=TRUE, eval=FALSE} 
map_dbl(.x = list("a", 10, 100),
        .f = possibly(log, NA))

## [1]       NA 2.302585 4.605170
```

---

# Glue 

- `glue` is a really cool package to concatenate things in a more readable manner 

- No more need for `paste()` and `paste0()` functions that look awuflly formatted 

- `glue` works on strings, SQL queries, data inside data frames etc

```{r, echo=TRUE, eval=FALSE}
name <- "Kaushik"
glue('My name is {name}.')
#> My name is Kaushik
```

- `glue_sql()` helps construct safe SQL queries, based on your DB flavour 

- My favourtie aspect of `glue_sql()` is it can expand vectors in an `IN` clause

```{r, echo=TRUE, eval=FALSE}
glue_sql(
  "SELECT * FROM my_table WHERE column_name IN ({vals*})",
  vals = c("filter1", "filter2"), .con = mysql_connection)
#> <SQL> SELECT * FROM my_table WHERE column_name IN ('filter1', 'filter2')
```


---

# Keyring

- Credential management is a huge problem in teams and organisations where code gets shared, especially on a web platform like github

- Scripts should not have passwords written into them 

- You can work around by storing passwords in hidden config files, but the `keyring` package offers a cleaner solution offering a variety of solutions like OS' credential store

- Use `keyring::key_set()` to set keys one time

- Then use `keyring::key_get()` to retrieve values in any script 

- Keyring offers 3 levels of depth - keyring, service and user

- Other functions include `keyring::key_list()`, `keyring::keyring_lock()`, `keyring::key_unlock()` etc

- Helps with reproducibility across users as well as systems

---

# Recipes

- During the lifecycle of building a predictive model, features need to be transformed 
  
- These transformation needs to be applied on the scoring/prediction set as well, which is a problem of reproducibility 

- The `recipes` package allows you to specify these transformations as steps of a recipe 


```{r, eval=FALSE, echo=TRUE}
model_recipe <- 
  recipe(Status ~ ., data = credit_train) %>%
  step_knnimpute(Home, Job, Marital, Income, Assets, Debt) %>%
  step_dummy(all_nominal(), -Status) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors()) %>%
  prep(training = credit_train)
```

- `prep` a recipe of steps on a training set

- `bake` the recipe on any dataset using a saved prepared recipe

```{r, eval=FALSE, echo=TRUE}
features_to_model <- bake(model_recipe,
                          new_data = credit_test, all_predictors())
```

---

# Other 

- `logger` for meaningful and hierarchical log messages

- `renv` and `packrat` help with capturing package and environment set up to make it reproducible 

- `workflows` an `drake` are packages built to help effective workflow mangaement, project management, reproducibility, and collaboration 

- Outside of the R world tools like `docker` and `git` help with taking reproducibility to the next level as well

---

# Conclusion

- Write code keeping in mind others, or future versions of you 

- Write code to be lazy & not have to rewrite stuff later 

- Write code that makes you feel like you are improving 

- Write code that sparks joy 

.center[
<img src = "imgs/marie_kondo_2.png" height=330>
]


---

class: center, middle

# Thanks!

Slides available on <a href = "https://github.com/kaushiklakshman/MelbURN-betteR"> my github </a>

For any questions, hit me up at `kaushiklakshman@gmail.com`